The association of smoking status with SARS-CoV-2 infection, hospitalisation and mortality from COVID-19: A living rapid evidence review with Bayesian meta-analyses (`r paste(current_version)`)

### 2021-07-27 reports/manuscript.Rmd
v12

---
title: 'The association of smoking status with SARS-CoV-2 infection, hospitalisation and mortality from COVID-19: A living rapid evidence review with Bayesian meta-analyses (`r paste(current_version)`)'
author: "David Simons, Lion Shahab, Jamie Brown and Olga Perski"
date: "`r format(Sys.time(), '%A %d %B %Y')`"
output:
  word_document:
    toc: yes
  html_document:
    highlight: zenburn
    toc: yes
    toc_float: yes
    number_sections: no
    theme: simplex
    css: C:/Users/ucbtds4/R_Repositories/smoking_covid_bayes/css/style.css
    keep_md: no
editor_options:
  chunk_output_type: inline
bibliography: manuscript.bib
csl: apa.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      warning = F,
                      message = F)
options(width = 180)
unlink('reports/markdown_report_cache', recursive = TRUE)
```

@@

**Findings:** `r paste(current_version)` (searches up to `r paste(tail(search_details$date_screened, 1))`) included `r paste(current_total)` studies with `r paste(nrow(running_meta_count))` ‘good’ and ‘fair’ quality studies included in unadjusted meta-analyses. `r paste(nrow(full_smoking_status))` studies (`r paste(round(nrow(full_smoking_status)/current_total*100,1), "%", sep = "")`) reported current, former and never smoking status with the remainder using broader categories. Recorded smoking prevalence among people with COVID-19 was generally lower than national prevalence. Current compared with never smokers were at reduced risk of SARS-CoV-2 infection (RR = `r paste(round(current_testing_RR_value, 2))`, 95% Credible Interval (CrI) = `r paste(round(current_testing_lci, 2), round(current_testing_uci, 2), sep = "-")`, τ = `r paste(round(current_testing_tau, 2))`). Data for former smokers were inconclusive (RR = `r paste(round(former_testing_RR_value, 2))`, 95% CrI = `r paste(round(former_testing_lci, 2), round(former_testing_uci, 2), sep = "-")`, τ = `r paste(round(former_testing_tau, 2))`) but favoured there being no important association (`r paste(round(1-m2_a_ecdf(1.1),2)*100, "%", sep = "")` probability of RR ≥1.1). Former compared with never smokers were at increased risk of hospitalisation (RR = `r paste(round(former_hospitalisation_RR_value, 2))`, CrI = `r paste(round(former_hospitalisation_lci, 2), round(former_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(former_hospitalisation_tau, 2))`), greater disease severity (RR = `r paste(round(former_severity_RR_value, 2))`, CrI = `r paste(round(former_severity_lci, 2), round(former_severity_uci, 2), sep = "-")`, τ = `r paste(round(former_severity_tau, 2))`) and mortality (RR = `r paste(round(former_mortality_RR_value, 2))`, CrI = `r paste(round(former_mortality_lci, 2), round(former_mortality_uci, 2), sep = "-")`, τ = `r paste(round(former_mortality_tau, 2))`). Data for current smokers on hospitalisation, disease severity and mortality were inconclusive (RR = `r paste(round(current_hospitalisation_RR_value, 2))`, 95% CrI = `r paste(round(current_hospitalisation_lci, 2), round(current_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(current_hospitalisation_tau, 2))`; RR `r paste(round(current_severity_RR_value, 2))`, 95% CrI = `r paste(round(current_severity_lci, 2), round(current_severity_uci, 2), sep = "-")`, τ = `r paste(round(current_severity_tau, 2))`; RR = `r paste(round(current_mortality_RR_value, 2))`, 95% CrI = `r paste(round(current_mortality_lci, 2), round(current_mortality_uci, 2), sep = "-")`, τ = `r paste(round(current_mortality_tau, 2))`, respectively) but favoured there being no important associations with hospitalisation and mortality (`r paste(round(1-m3_a_ecdf(1.1),2)*100, "%", sep = "")` and `r paste(round(1-m7_a_ecdf(1.1),2)*100, "%", sep = "")` probability of RR ≥1.1, respectively) and a small but important association with disease severity (`r paste(round(1-m5_a_ecdf(1.1),2)*100, "%", sep = "")` probability of RR ≥1.1).

**Conclusions:** Compared with never smokers, current smokers appear to be at reduced risk of SARS-CoV-2 infection while former smokers appear to be at increased risk of hospitalisation, greater disease severity and mortality from COVID-19. However, it is uncertain whether these associations are causal.

@@

Data were extracted by one reviewer and verified (i.e. independently checked against pre-prints and published reports) by another on i) author (year); ii) date published; iii) country; iv) study design; v) study setting; vi) sample size; vii) sex; viii) age; ix) smoking status (e.g. current, former, never, not stated, missing) and whether it was biochemically verified; x) use of alternative nicotine products; xi) SARS-CoV-2 testing; xii) SARS-CoV-2 infection; xiii) diagnosis of COVID-19; xiv) hospitalisation with COVID-19; xv) disease severity in those hospitalised with COVID-19; xvi) mortality; xvii) adjustment of smoking specific risk estimates for relevant covariates (e.g. age, sex); and xviii) whether a representative or random sampling method was used.

@@

A narrative synthesis was conducted. Data from ‘good’ and ‘fair’ quality studies were pooled in R v.3.6.3 [@team_r_2013]. In a living review where new data are regularly added to the analyses, it may be more appropriate to use a Bayesian (as opposed to frequentist) approach where prior knowledge is used in combination with new data to estimate a posterior risk distribution. A Bayesian approach mitigates the issue of performing multiple statistical tests, which can inflate family-wise error. A series of random-effects hierarchical Bayesian meta-analyses were performed with the brms [@burkner_advanced_2018] package to estimate the relative risk for each comparison with accompanying 95% credible intervals (CrIs). We first defined prior distributions for the true pooled effect size (µ) and the between-study heterogeneity (τ), with µ specified as a normal distribution with a mean equal to the derived point estimate from each comparison of interest in the immediately preceding version of this living review, and τ specified as a half-Cauchy distribution with a mean of 0 and standard deviation of 1. The half-Cauchy distribution was selected to reflect prior knowledge that high levels of between-study heterogeneity are more likely than lower levels. Markov Chain Monte Carlo methods (20,000 burn-ins followed by 80,000 iterations) were then used to generate a risk distribution for each study, in addition to a pooled effect for the posterior risk distribution. We report forest plots with the pooled effect for the posterior risk distribution displayed as the median relative risk with an accompanying 95% CrIs. We used the empirical cumulative distribution function (ECDF) to estimate the probability of there being a 10% reduction or 10% increase in relative risk (RR) (i.e. RR ≥1.1 or RR ≤0.9). Due to a lack of indication as to what constitutes a clinically or epidemiologically meaningful effect (e.g. with regards to onward disease transmission or requirements for intensive care beds), we deemed a 10% change in risk as small but important. Where data were inconclusive (as indicated by CrIs crossing RR = 1.0), to disambiguate whether data favoured no effect or there being a small but important association, we estimated whether there was ≥75% probability of RR ≥1.1 or RR ≤0.9.

@@

In the current review version (`r paste(current_version)`) with searches up to `r paste(tail(search_details$date_screened, 1))`, a total of `r paste(prisma$"Titles/abstracts screened")` records were identified, with `r paste(prisma$"Studies in narrative synthesis")` studies included in a narrative synthesis and `r paste(nrow(running_meta_count))` studies included in meta-analyses (see [Figure 1](#fig-1)).

@@

Characteristics of included studies are presented in [Table 1](#table-1). Studies were conducted across `r paste(nrow(country))` countries. `r paste(all_countries[[1]][1,2])` studies were conducted in the `r paste(all_countries[[1]][1,1])`, `r paste(all_countries[[2]][1,2])` in `r paste(all_countries[[2]][1,1])`, `r paste(all_countries[[3]][1,2])` in the `r paste(all_countries[[3]][1,1])`, `r paste(all_countries[[4]][1,2])`  in `r paste(all_countries[[4]][1,1])`, `r paste(all_countries[[5]][1,2])`  in `r paste(all_countries[[5]][1,1])`, `r paste(all_countries[[6]][1,2])` in `r paste(all_countries[[6]][1,1])`, `r paste(all_countries[[7]][1,2])` in `r paste(all_countries[[7]][1,1])`, `r paste(all_countries[[8]][1,2])` in `r paste(all_countries[[8]][1,1])`, `r paste(all_countries[[9]][1,2])` in `r knitr::combine_words(unlist(all_countries[[9]][1]))`, `r paste(all_countries[[10]][1,2])` in `r knitr::combine_words(unlist(all_countries[[10]][1]))`, `r paste(all_countries[[11]][1,2])` in `r knitr::combine_words(unlist(all_countries[[11]][1]))`, `r paste(all_countries[[12]][1,2])` in `r knitr::combine_words(unlist(all_countries[[12]][1]))`, with `r paste(all_countries[[13]][1,2])` each from `r knitr::combine_words(unlist(all_countries[[13]][1]))` and a single study from `r paste(nrow(all_countries[[14]][1]))` further countries. The majority of studies used observational designs (see [Supplementary table S1](#supplementary-t1)). `r paste(setting[1,2])` (`r paste(round(setting[1,2]/sum(setting[2])*100, 1))`%) were conducted in `r paste(tolower(setting[1,1]))` settings, `r paste(setting[2,2])` studies (`r paste(round(setting[2,2]/sum(setting[2])*100, 1))`%) included individuals from `r paste(tolower(setting[2,1]))` settings, `r paste(setting[3,2])` studies (`r paste(round(setting[3,2]/sum(setting[2])*100, 1))`%) were conducted exclusively in the `r paste(tolower(setting[3,1]))`, with one study each conducted in a homeless shelter and a quarantine centre, and one study that did not state the study setting. Studies had a median of `r paste(participants[3])` (interquartile range = `r paste(participants[2], participants[5], sep = "-")`) participants. The majority of studies (`r paste(round(sum(diagnostics[5]+diagnostics[6])/sum(diagnostics)*100,1))`%) used reverse transcriptase polymerase chain reaction (RT-PCR) for confirmation of SARS-CoV-2 infection, `r paste(round(sum(diagnostics[1]+diagnostics[6])/sum(diagnostics)*100,1))`% used an antibody test to confirm prior infection and `r paste(round(diagnostics[6]/sum(diagnostics)*100,1), "%", sep = "")` of studies relied on a combination of RT-PCR or antibody assays.

@@

Categorisation of smoking status was heterogeneous (see [Table 1](#table-1)). `r paste(data_source[1,2])` studies collected data on smoking status through routine electronic health records (EHRs), `r paste(data_source[2,2])` studies used a bespoke case report form for COVID-19 and `r paste(data_source[3,2])` studies did not state the source for information on smoking status. None of the studies verified smoking status biochemically. Notably, only `r paste(nrow(full_smoking_status))` (`r paste(round(nrow(full_smoking_status)/current_total*100,1))`%) studies reported current, former and never smoking status (see [Supplementary table S2a](#supplementary-t2a)), with a further 26 studies reporting only ever and never smoking status (see [Supplementary table S2b](#supplementary-t2b)). The remaining `r paste(nrow(incomplete_smoking_status))` studies reported current, current/former or current and former smoking status but did not explicitly state whether remaining participants were never smokers or if data were missing on smoking status (see [Supplementary table S2c](#supplementary-t2c)). `r paste(nrow(missing_smok))` studies explicitly reported the proportion with missing data on smoking status, which ranged from `r paste(round(minimum,1), "%", sep = "")` to `r paste(round(maximum,1), "%", sep = "")`.

@@

Eight studies recorded the use of alternative nicotine products in current and/or former smokers but did not report COVID-19 outcomes stratified by alternative nicotine use [@miyara_low_2020; @rimland_clinical_2020; @girardeau_confirmed_2020; @ebinger_sars-cov-2_2020; @islam_risk_2020; @gallichotte_longitudinal_2020; @kantele_sars-cov-2_2020; @crooks_predicting_2020]. One additional study grouped together current smokers and vapers [@schubl_risk_2020].

@@

Three studies were performed in random or representative population samples and were rated as ‘good’ quality, and `r paste(nrow(fair_quality_number))` studies were rated as ‘fair’ quality, of which `r paste(nrow(running_meta_count))` studies reported results stratified by smoking status for the outcomes of interest and could be included in meta-analyses. The remaining `r paste(nrow(poor_quality_number))` studies were rated as ‘poor’ quality (see [Table 1](#table-1)).

@@

set_flextable_defaults(
  font.family = "Calibri", 
  font.size = 8,
  font.color = "black",
  digits = 2,
  theme_fun = "theme_booktabs"
  )

tbl_1 <- readRDS(file = here::here("reports", "tables", "table_1_refs.rds")) %>%
  arrange(`Date published`) %>%
  mutate(`Lead author` = recode(`Lead author`, "Schubl" = "Schuble*"))

flextable(tbl_1) %>%
  colformat_md(j = "ref") %>%
  colformat_double(j = "Sample size", digits = 0)
```

@@

![*Figure 2a.* Weighted mean prevalence of current smoking in included studies, split by country (solid lines). The circles represent individual studies, with colour corresponding to study setting (i.e. community, community and hospital, hospital) and size corresponding to relative study sample size. For comparison, national current smoking prevalence is shown by the dashed red lines. Countries with three or more eligible studies are shown.](figure/current_smoking_plots_updated.png){#fig-2a}

![*Figure 2b.* Prevalence of former smoking in included studies, split by country (solid lines). The circles represent individual studies, with colour corresponding to study setting (i.e. community, community and hospital, hospital) and size corresponding to relative study sample size. For comparison, national former smoking prevalence is shown by dashed red lines. Countries with three or more eligible studies are shown.](figure/former_smoking_plots_updated.png){#fig-2b}

@@

`r paste(nrow(table_2_word))` studies provided data on SARS-CoV-2 infection for people meeting local testing criteria by smoking status (see [Table 2](#table-2)). Meta-analyses were performed for `r paste(nrow(good_quality_2))` ‘good’ and `r paste(nrow(fair_quality_2))` ‘fair’ quality studies (see [Figure 3](#fig-3) and [4](#fig-4)). Current smokers were at reduced risk of testing positive for SARS-CoV-2 compared with never smokers (RR = `r paste(round(current_testing_RR_value, 2))`, 95% Credible Interval (CrI) = `r paste(round(current_testing_lci, 2), round(current_testing_uci, 2), sep = "-")`, τ = `r paste(round(current_testing_tau, 2))`). The three good quality studies each reported point estimates less than 1, although the CrI for one of the three studies crossed 1. The probability of current smokers being at reduced risk of infection compared with never smokers (RR ≤0.9) was >99%. Former compared with never smokers were at increased risk of testing positive, but data were inconclusive (RR = `r paste(round(former_testing_RR_value, 2))`, 95% CrI = `r paste(round(former_testing_lci, 2), round(former_testing_uci, 2), sep = "-")`, τ = `r paste(round(former_testing_tau, 2))`) and favoured there being no important association. The probability of former smokers being at increased risk of infection (RR ≥1.1) compared with never smokers was `r paste(round(1-m2_a_ecdf(1.1), 2)*100, "%", sep = "")`. Results were materially unchanged in sensitivity analyses.

@@

![*Figure 3.* Forest plot for risk of testing positive for SARS-CoV-2 in current vs. never smokers. * Indicates ‘good’ quality studies. The prior from the previous review version (v10) was RR = 0.69.](figure/m1_a.png){#fig-3}

![*Figure 4.* Forest plot for risk of testing positive for SARS-CoV-2 in former vs. never smokers. * Indicates ‘good’ quality studies. The prior from the previous review version (v10) was RR = 1.02.](figure/m2_a.png){#fig-4}

@@

`r paste(nrow(table_3_word))` studies examined hospitalisation for COVID-19 disease, stratified by smoking status (see [Table 3](#table-3)). Meta-analyses were performed for `r paste(nrow(fair_quality_3))` ‘fair’ quality studies (see [Figure 5](#fig-5) and [6](#fig-6)). Current (RR = `r paste(round(current_hospitalisation_RR_value, 2))`, 95% CrI = `r paste(round(current_hospitalisation_lci, 2), round(current_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(current_hospitalisation_tau, 2))`) and former (RR = `r paste(round(former_hospitalisation_RR_value, 2))`, CrI = `r paste(round(former_hospitalisation_lci, 2), round(former_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(former_hospitalisation_tau, 2))`) compared with never smokers were at increased risk of hospitalisation with COVID-19. However, data for current smokers were inconclusive and favoured there being no important association. The probability of current and former smokers being at increased risk of hospitalisation (RR ≥1.1) compared with never smokers was `r paste(round(1-m3_a_ecdf(1.1),2)*100,"%", sep = "")` and `r paste(round(1-m4_a_ecdf(1.1),2)*100, "%", sep = "")`, respectively. Results were materially unchanged in two sensitivity analyses.

@@

![*Figure 5.* Forest plot for risk of hospitalisation in current vs. never smokers. The prior from the previous review version (v10) was RR = 1.08.](figure/m3_a.png){#fig-5}

![*Figure 6.* Forest plot for risk of hospitalisation in former vs. never smokers. The prior from the previous version (v10) was RR = 1.18.](figure/m4_a.png){#fig-5}

@@

`r paste(nrow(table_4_word))` studies reported disease severity in hospitalised patients stratified by smoking status (see [Table 4](#table-4)). Severe (as opposed to non-severe) disease was broadly defined as requiring intensive treatment unit (ITU) admission, requiring oxygen as a hospital inpatient or in-hospital death. Meta-analyses were performed for `r paste(nrow(fair_quality_4))` ‘fair’ quality studies (see [Figure 7](#fig-7) and [8](#fig-8)). Current (RR = `r paste(round(current_severity_RR_value, 2))`, 95% CrI = `r paste(round(current_severity_lci, 2), round(current_severity_uci, 2), sep = "-")`, τ = `r paste(round(current_severity_tau, 2))`) and former (RR = `r paste(round(former_severity_RR_value, 2))`, 95% CrI = `r paste(round(former_severity_lci, 2), round(former_severity_uci, 2), sep = "-")`, τ = `r paste(round(former_severity_tau, 2))`) compared with never smokers were at increased risk of greater disease severity. However, while data for current smokers only were inconclusive, they favoured there being a small but important association. The probability of current and former smokers having increased risk of greater disease severity (RR ≥1.1) compared with never smokers was `r paste(round(1-m5_a_ecdf(1.1),2)*100, "%", sep = "")` and `r paste(round(1-m6_a_ecdf(1.1),2)*100, "%", sep = "")`, respectively. Results were materially unchanged in two sensitivity analyses.

>>>>

---
title: 'The association of smoking status with SARS-CoV-2 infection, hospitalisation and mortality from COVID-19: A living rapid evidence review with Bayesian meta-analyses (`r paste(current_version)`)'
author: "David Simons, Lion Shahab, Jamie Brown and Olga Perski"
date: "`r format(Sys.time(), '%A %d %B %Y')`"
output:
  word_document:
    toc: no
  html_document:
    highlight: zenburn
    number_sections: no
    theme: simplex
    keep_md: no
editor_options:
  chunk_output_type: inline
bibliography: manuscript.bib
csl: apa.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      warning = F,
                      message = F)
options(width = 180)
unlink('reports/markdown_report_cache', recursive = TRUE)
```

@@

**Findings:** `r paste(current_version)` (searches up to `r paste(tail(search_details$date_screened, 1))`) included `r paste(current_total)` studies with `r paste(nrow(running_meta_count))` ‘good’ and ‘fair’ quality studies included in unadjusted meta-analyses. `r paste(nrow(full_smoking_status))` studies (`r paste(round(nrow(full_smoking_status)/current_total*100,1), "%", sep = "")`) reported current, former and never smoking status with the remainder using broader categories. Recorded smoking prevalence among people with COVID-19 was generally lower than national prevalence. Current compared with never smokers were at reduced risk of SARS-CoV-2 infection (RR = `r paste(round(current_testing_RR_value, 2))`, 95% Credible Interval (CrI) = `r paste(round(current_testing_lci, 2), round(current_testing_uci, 2), sep = "-")`, τ = `r paste(round(current_testing_tau, 2))`). Data for former smokers were inconclusive (RR = `r paste(round(former_testing_RR_value, 2))`, 95% CrI = `r paste(round(former_testing_lci, 2), round(former_testing_uci, 2), sep = "-")`, τ = `r paste(round(former_testing_tau, 2))`) but favoured there being no important association (`r paste(round(1-m2_a_ecdf(1.1),2)*100, "%", sep = "")` probability of RR ≥1.1). Former compared with never smokers were at increased risk of hospitalisation (RR = `r paste(round(former_hospitalisation_RR_value, 2))`, CrI = `r paste(round(former_hospitalisation_lci, 2), round(former_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(former_hospitalisation_tau, 2))`), greater disease severity (RR = `r paste(round(former_severity_RR_value, 2))`, CrI = `r paste(round(former_severity_lci, 2), round(former_severity_uci, 2), sep = "-")`, τ = `r paste(round(former_severity_tau, 2))`) and mortality (RR = `r paste(round(former_mortality_RR_value, 2))`, CrI = `r paste(round(former_mortality_lci, 2), round(former_mortality_uci, 2), sep = "-")`, τ = `r paste(round(former_mortality_tau, 2))`). Current compared with never smokers were at increased risk of greater disease severity (RR `r paste(round(current_severity_RR_value, 2))`, 95% CrI = `r paste(round(current_severity_lci, 2), round(current_severity_uci, 2), sep = "-")`, τ = `r paste(round(current_severity_tau, 2))`). Data for current smokers on hospitalisation and mortality were inconclusive (RR = `r paste(round(current_hospitalisation_RR_value, 2))`, 95% CrI = `r paste(round(current_hospitalisation_lci, 2), round(current_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(current_hospitalisation_tau, 2))`; RR = `r paste(round(current_mortality_RR_value, 2))`, 95% CrI = `r paste(round(current_mortality_lci, 2), round(current_mortality_uci, 2), sep = "-")`, τ = `r paste(round(current_mortality_tau, 2))`, respectively) but favoured there being no important associations (`r paste(round(1-m3_a_ecdf(1.1),2)*100, "%", sep = "")` and `r paste(round(1-m7_a_ecdf(1.1),2)*100, "%", sep = "")` probability of RR ≥1.1, respectively).

**Conclusions:** Compared with never smokers, current smokers appear to be at reduced risk of SARS-CoV-2 infection and increased risk of greater disease severity, while former smokers appear to be at increased risk of hospitalisation, greater disease severity and mortality from COVID-19. However, it is uncertain whether these associations are causal.

@@

Data were extracted by one reviewer and verified (i.e. independently checked against pre-prints and published reports) by another on i) author (year); ii) date published; iii) country; iv) study design; v) study setting; vi) sample size; vii) sex; viii) age; ix) smoking status (e.g. current, former, never, not stated, missing) and whether it was biochemically verified; x) use of alternative nicotine products (e.g. e-cigarettes); xi) SARS-CoV-2 testing; xii) SARS-CoV-2 infection; xiii) diagnosis of COVID-19; xiv) hospitalisation with COVID-19; xv) disease severity in those hospitalised with COVID-19; xvi) mortality; xvii) adjustment of smoking specific risk estimates for relevant covariates (e.g. age, sex); and xviii) whether a representative or random sampling method was used.

@@

A narrative synthesis was conducted. Data from ‘good’ and ‘fair’ quality studies were pooled in R v.3.6.3 [@team_r_2013]. In a living review where new data are regularly added to the analyses, it may be more appropriate to use a Bayesian (as opposed to frequentist) approach where prior knowledge is used in combination with new data to estimate a posterior risk distribution. A Bayesian approach mitigates the issue of performing multiple statistical tests, which can inflate family-wise error. A series of random-effects hierarchical Bayesian meta-analyses were performed with the brms [@burkner_advanced_2017] package to estimate the relative risk for each comparison with accompanying 95% credible intervals (CrIs). We first defined prior distributions for the true pooled effect size (µ) and the between-study heterogeneity (τ), with µ specified as a normal distribution with a mean equal to the derived point estimate from each comparison of interest in the immediately preceding version of this living review, and τ specified as a half-Cauchy distribution with a mean of 0 and standard deviation of 1. The half-Cauchy distribution was selected to reflect prior knowledge that high levels of between-study heterogeneity are more likely than lower levels. Markov Chain Monte Carlo methods (20,000 burn-ins followed by 80,000 iterations) were then used to generate a risk distribution for each study, in addition to a pooled effect for the posterior risk distribution. We report forest plots with the pooled effect for the posterior risk distribution displayed as the median relative risk with an accompanying 95% CrIs. We used the empirical cumulative distribution function (ECDF) to estimate the probability of there being a 10% reduction or 10% increase in relative risk (RR) (i.e. RR ≥1.1 or RR ≤0.9). Due to a lack of indication as to what constitutes a clinically or epidemiologically meaningful effect (e.g. with regards to onward disease transmission or requirements for intensive care beds), we deemed a 10% change in risk as small but important. Where data were inconclusive (as indicated by CrIs crossing RR = 1.0), to disambiguate whether data favoured no effect or there being a small but important association, we estimated whether there was ≥75% probability of RR ≥1.1 or RR ≤0.9.

@@

In the current review version (`r paste(current_version)`) with searches up to `r paste(tail(search_details$date_screened, 1))`, a total of 1940 records were identified, with `r paste(prisma$"Studies in narrative synthesis")` studies included in a narrative synthesis and `r paste(nrow(running_meta_count))` studies included in meta-analyses (see [Figure 1](#fig-1)).

@@

Characteristics of included studies are presented in [Table 1](#table-1). Studies were conducted across `r paste(nrow(country))` countries. `r paste(all_countries[[1]][1,2])` studies were conducted in the USA, `r paste(all_countries[[2]][1,2])` in China, `r paste(all_countries[[3]][1,2])` in the UK, `r paste(all_countries[[4]][1,2])`  in Spain, `r paste(all_countries[[5]][1,2])`  in  Italy, `r paste(all_countries[[6]][1,2])` in  France, `r paste(all_countries[[7]][1,2])` in Mexico, `r paste(all_countries[[8]][1,2])` in multiple countries, `r paste(all_countries[[9]][1,2])` in Switzerland, `r paste(all_countries[[10]][1,2])` in `r knitr::combine_words(unlist(all_countries[[10]][1]))`, `r paste(all_countries[[11]][1,2])` in `r knitr::combine_words(unlist(all_countries[[11]][1]))`, `r paste(all_countries[[12]][1,2])` in `r knitr::combine_words(unlist(all_countries[[12]][1]))`, with `r paste(all_countries[[13]][1,2])` in `r knitr::combine_words(unlist(all_countries[[13]][1]))` and 97 studies from 39 further countries. The majority of studies used observational designs (see [Supplementary table S1](#supplementary-t1)). `r paste(setting[1,2])` (`r paste(round(setting[1,2]/sum(setting[2])*100, 1))`%) were conducted in `r paste(tolower(setting[1,1]))` settings, `r paste(setting[2,2])` studies (`r paste(round(setting[2,2]/sum(setting[2])*100, 1))`%) included individuals from `r paste(tolower(setting[2,1]))` settings, `r paste(setting[3,2])` studies (`r paste(round(setting[3,2]/sum(setting[2])*100, 1))`%) were conducted exclusively in the `r paste(tolower(setting[3,1]))`, with one study each conducted in a homeless shelter and a quarantine centre, and one study that did not state the study setting. Studies had a median of `r paste(participants[3])` (interquartile range = `r paste(round(participants[2], 0), round(participants[5], 0), sep = "-")`) participants. The majority of studies (`r paste(round(sum(diagnostics[6]+diagnostics[7])/sum(diagnostics)*100,1))`%) used reverse transcriptase polymerase chain reaction (RT-PCR) for confirmation of SARS-CoV-2 infection, `r paste(round(sum(diagnostics[1]+diagnostics[7])/sum(diagnostics)*100,1))`% used an antibody test to confirm prior infection and `r paste(round(sum(diagnostics[7] + diagnostics[2])/sum(diagnostics)*100,1), "%", sep = "")` of studies relied on a combination of RT-PCR, antigen or antibody assays.

@@

Categorisation of smoking status was heterogeneous (see [Table 1](#table-1)). `r paste(data_source[1,2])` `r paste(round(data_source[1,2]/sum(data_source[,2])*100,1), "%", sep = "")` studies collected data on smoking status through routine electronic health records (EHRs), `r paste(data_source[2,2])` `r paste(round(data_source[2,2]/sum(data_source[,2])*100,1), "%", sep = "")` studies used a bespoke case report form for COVID-19 and `r paste(data_source[3,2])` `r paste(round(data_source[3,2]/sum(data_source[,2])*100,1), "%", sep = "")` studies did not state the source for information on smoking status. None of the studies verified smoking status biochemically. Notably, only `r paste(nrow(full_smoking_status))` (`r paste(round(nrow(full_smoking_status)/current_total*100,1))`%) studies reported current, former and never smoking status (see [Supplementary table S2a](#supplementary-t2a)), with a further 26 studies reporting only ever and never smoking status (see [Supplementary table S2b](#supplementary-t2b)). The remaining `r paste(nrow(incomplete_smoking_status))` studies reported current, current/former or current and former smoking status but did not explicitly state whether remaining participants were never smokers or if data were missing on smoking status (see [Supplementary table S2c](#supplementary-t2c)). `r paste(nrow(missing_smok))` studies explicitly reported the proportion with missing data on smoking status, which ranged from `r paste(round(minimum,1), "%", sep = "")` to `r paste(round(maximum,1), "%", sep = "")`.

@@

Fifteen studies recorded the use of alternative nicotine products in current and/or former smokers but did not report COVID-19 outcomes stratified by alternative nicotine use (see [Table 1](#table-1).

@@

Nine studies were performed in random or representative population samples and were rated as ‘good’ quality, and `r paste(nrow(quality_rating %>% filter(overall_quality == "fair")))` studies were rated as ‘fair’ quality, of which `r paste(nrow(running_meta_count))` studies reported results stratified by smoking status for the outcomes of interest and could be included in meta-analyses. The remaining `r paste(nrow(poor_quality_number))` studies were rated as ‘poor’ quality (see [Table 1](#table-1)).

@@

set_flextable_defaults(
  font.family = "Calibri", 
  font.size = 8,
  font.color = "black",
  digits = 1,
  theme_fun = "theme_booktabs"
  )

tbl_1 <- readRDS(file = here::here("reports", "tables", "table_1_refs.rds")) %>%
  arrange(`Date published`) %>%
  mutate(`Lead author` = recode(`Lead author`, "Schubl" = "Schuble*"),
         `Missing %` = as.numeric(`Missing %`))

flextable(tbl_1) %>%
  colformat_md(j = "ref") %>%
  colformat_double(j = "Sample size", digits = 0) %>%
  colformat_double(j = 8:21, digits = 1)
```

@@

![*Figure 2.* Weighted mean prevalence of current (left panel) and former (right panel) smoking in included studies, split by country. The circles represent individual studies, with colour corresponding to study setting (i.e. hospital, community and hospital, community) and size corresponding to relative study sample size. For comparison, national current and former smoking prevalence are indicated by the red lines. Countries with three or more eligible studies are shown.](figure/combined_smoking_plots_updated.png){#fig-2a}

@@

`r paste(nrow(table_2_word))` studies provided data on SARS-CoV-2 infection for people meeting local testing criteria by smoking status (see [Table 2](#table-2)). Meta-analyses were performed for `r paste(nrow(good_quality_2))` ‘good’ and `r paste(nrow(fair_quality_2))` ‘fair’ quality studies (see [Figure 3](#fig-3) and [4](#fig-4)). Current smokers were at reduced risk of testing positive for SARS-CoV-2 compared with never smokers (RR = `r paste(round(current_testing_RR_value, 2))`, 95% Credible Interval (CrI) = `r paste(round(current_testing_lci, 2), round(current_testing_uci, 2), sep = "-")`, τ = `r paste(round(current_testing_tau, 2))`). The six good quality studies each reported point estimates less than 1, although the CrI for two of the six studies crossed 1. The probability of current smokers being at reduced risk of infection compared with never smokers (RR ≤0.9) was >99%. Former compared with never smokers were at increased risk of testing positive, but data were inconclusive (RR = `r paste(round(former_testing_RR_value, 2))`, 95% CrI = `r paste(round(former_testing_lci, 2), round(former_testing_uci, 2), sep = "-")`, τ = `r paste(round(former_testing_tau, 2))`) and favoured there being no important association. The probability of former smokers being at increased risk of infection (RR ≥1.1) compared with never smokers was < 1%. Results were materially unchanged in sensitivity analyses.

@@

![*Figure 3.* Forest plot for risk of testing positive for SARS-CoV-2 in current vs. never smokers. * Indicates ‘good’ quality studies. The prior from the previous review version (v11) was RR = 0.71.](figure/m1_a.png){#fig-3}

![*Figure 4.* Forest plot for risk of testing positive for SARS-CoV-2 in former vs. never smokers. * Indicates ‘good’ quality studies. The prior from the previous review version (v11) was RR = 1.03.](figure/m2_a.png){#fig-4}

@@

`r paste(nrow(table_3_word))` studies examined hospitalisation for COVID-19 disease, stratified by smoking status (see [Table 3](#table-3)). Meta-analyses were performed for one 'good' and 18 ‘fair’ quality studies (see [Figure 5](#fig-5) and [6](#fig-6)). Current (RR = `r paste(round(current_hospitalisation_RR_value, 2))`, 95% CrI = `r paste(round(current_hospitalisation_lci, 2), round(current_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(current_hospitalisation_tau, 2))`) and former (RR = `r paste(round(former_hospitalisation_RR_value, 2))`, CrI = `r paste(round(former_hospitalisation_lci, 2), round(former_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(former_hospitalisation_tau, 2))`) compared with never smokers were at increased risk of hospitalisation with COVID-19. However, data for current smokers were inconclusive and favoured there being no important association. The probability of current and former smokers being at increased risk of hospitalisation (RR ≥1.1) compared with never smokers was `r paste(round(1-m3_a_ecdf(1.1),2)*100,"%", sep = "")` and >99%, respectively. Results were materially unchanged in two sensitivity analyses.

@@

![*Figure 5.* Forest plot for risk of hospitalisation in current vs. never smokers. The prior from the previous review version (v11) was RR = 1.1.](figure/m3_a.png){#fig-5}

![*Figure 6.* Forest plot for risk of hospitalisation in former vs. never smokers. The prior from the previous version (v11) was RR = 1.19.](figure/m4_a.png){#fig-5}

@@

`r paste(nrow(table_4_word))` studies reported disease severity in hospitalised patients stratified by smoking status (see [Table 4](#table-4)). Severe (as opposed to non-severe) disease was broadly defined as requiring intensive treatment unit (ITU) admission, requiring oxygen as a hospital inpatient or in-hospital death. Meta-analyses were performed for 16 ‘fair’ quality studies (see [Figure 7](#fig-7) and [8](#fig-8)). Current (RR = `r paste(round(current_severity_RR_value, 2))`, 95% CrI = `r paste(round(current_severity_lci, 2), round(current_severity_uci, 2), sep = "-")`, τ = `r paste(round(current_severity_tau, 2))`) and former (RR = `r paste(round(former_severity_RR_value, 2))`, 95% CrI = `r paste(round(former_severity_lci, 2), round(former_severity_uci, 2), sep = "-")`, τ = `r paste(round(former_severity_tau, 2))`) compared with never smokers were at increased risk of greater disease severity. The probability of current and former smokers having increased risk of greater disease severity (RR ≥1.1) compared with never smokers was `r paste(round(1-m5_a_ecdf(1.1),2)*100, "%", sep = "")` and `r paste(round(1-m6_a_ecdf(1.1),2)*100, "%", sep = "")`, respectively. Results were materially unchanged in two sensitivity analyses.
